---
title: "Week 4 Challenge Standard"
output: html_document
---

(1) Use `tableone` to create a basic table that includes the appropriate descriptive statistics for following variables from the NHANES Drug Use 2013-2014 questionnaire: DUQ200, DUQ210, DUQ240, DUQ260. 

Before creating the table: 

* recode the variables so they have clear variable names, clear category labels, and values that don't make sense are coded to `NA`  
* check the distribution to decide which descriptive statistics to report for any continuous variables 

```{r}
# bring in data
library(RNHANES)
nhanes2013 <- nhanes_load_data(file_name = "DUQ_H",
                               year = "2013-2014",
                               demographics = TRUE)

# open library for recoding 
library(car)

# recode DUQ200
nhanes2013$marijuanaUse <- recode(nhanes2013$DUQ200,
                                  "1 = 'Yes';
                                  2 = 'No'; 7 = NA; 9 = NA")

# recode DUQ210
nhanes2013$age.first.mari <- recode(nhanes2013$DUQ210,
                                  "6 = NA;
                                  777 = NA; 999 = NA")

# recode DUQ240 
nhanes2013$cocaineUse <- recode(nhanes2013$DUQ240,
                                  "1 = 'Yes';
                                  2 = 'No'; 7 = NA; 9 = NA")

# recode DUQ260
nhanes2013$age.first.cocaine <- recode(nhanes2013$DUQ260,
                                  "6 = NA;
                                  777 = NA; 999 = NA")
```

```{r}
# check distributions for continuous variables 
library(ggplot2)
ggplot(data = nhanes2013, aes(x = age.first.mari)) +
  geom_histogram()
ggplot(data = nhanes2013, aes(x = age.first.cocaine)) +
  geom_histogram()

```

```{r}
# create table using tableone 
# open library
library(tableone)

# create table 
drug.table <- CreateTableOne(vars = c("age.first.mari", "age.first.cocaine", 
                                      'marijuanaUse','cocaineUse'), data = nhanes2013)

# print out table basic
print(drug.table, nonnormal = c('age.first.cocaine'),
      showAllLevels = TRUE)

```

(2) Use `kableExtra` to create a fancy table that includes the appropriate descriptive statistics for the same four variables from question 1. Format the table so that all the information is needed to understand the values in the table. 

```{r}
# open the library
library(kableExtra)

# create the table with formatting 
# get descriptives for marijuana use
marijuana.use <- cbind(table(nhanes2013$marijuanaUse),
                    round(100*(prop.table(table(nhanes2013$marijuanaUse))), 1))

# add column names
colnames(marijuana.use) <- c("Frequency", "Percentage")

# make data frame
marijuana.use <- data.frame(marijuana.use)

# get descriptives for meth use
cocaine.use <- cbind(table(nhanes2013$cocaineUse),
                    round(100*(prop.table(table(nhanes2013$cocaineUse))), 1))

# add column names
colnames(cocaine.use) <- c("Frequency", "Percentage")

# make data frame
cocaine.use <- data.frame(cocaine.use)

# combine data frames 
drug.use <- rbind(marijuana.use, cocaine.use)

# move row labels to a variable named Used
drug.use <- data.frame(Used = row.names(drug.use), drug.use) 

# replace row labels with accurate labels
rownames(drug.use) <- c("Marijuana.no", "Marijuana.yes", "Cocaine.no", "Cocaine.yes")

# recode No1 and Yes1 to No and Yes
drug.use$Used <- recode(drug.use$Used, "'No1' = 'No'; 'Yes1' = 'Yes'")

# change drug.use Used variable to string
drug.use$Used <- as.character(drug.use$Used)

# make vectors of the median and IQR for
# age meth first use and age heroin first use
age.mari.desc <- c("Marijuana (mean, sd)", round(mean(nhanes2013$age.first.mari, na.rm = T),1),
                  round(sd(nhanes2013$age.first.mari, na.rm = T), 1))

age.cocaine.desc <- c("Cocaine (median, IQR)", median(nhanes2013$age.first.cocaine, na.rm = T),
                  IQR(nhanes2013$age.first.cocaine, na.rm = T))

# add vectors as rows to the drug.use data frame
drug.use.all <- rbind(drug.use, age.mari.desc, age.cocaine.desc)

# print the data frame to check it
drug.use.all

# open kableExtra package
library(kableExtra)
  
# table with all four variables
kable(drug.use.all, "html", caption = "Drug use among NHANES 2013-2014 participants",
      row.names = FALSE) %>%
  kable_styling("striped", full_width = TRUE) %>%
  group_rows("Ever used marijuana", 1, 2) %>%
  group_rows("Ever used cocaine, crack, heroin, or meth", 3, 4) %>% 
  group_rows("Age in years of first use", 5, 6) 
```

(3) Use `waffle` to create well-formatted waffle plots for the two categorical variables from question 1. Find colors on colorbrewer to distinguish the groups.

```{r}
# open the library
library(waffle)


# create the plots 
waffle(table(nhanes2013$marijuanaUse)/50, rows=5,
       title = "NHANES participants who ever used marijuana, 2013-2014",
       colors = c("#af8dc3", "#7fbf7b", "white"))
waffle(table(nhanes2013$cocaineUse)/50, rows=5,
       title = "NHANES participants who ever used cocaine, crack,\nheroin, or meth, 2013-2014",
       colors = c("#af8dc3", "#7fbf7b", "white"))

```

(4) Write a paragraph intepreting the information in the tables and waffle plot.

More people have tried marijuana (54%) than cocaine, crack, heroin, or meth (16%). The mean age of first marijuana use was 17.1 (sd = 4.5), which was younger than the median age of first use of cocaine, crack, heroin, or meth (med = 20; IQR = 6).
