---
title: "Challenge 10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Complete ONE of the following:

* DataCamp Multiple and Logistic Regression Course Chapter 4 (If you do this, please submit a note in Blackboard that you completed the DataCamp version of the challenge)

OR
 
* Usual style of challenge

### Standard edition

Use the University of Kentucky data set (codebook below) and logistic regression to answer this question: _How well do age of initiation, percent of friends who smoke, and motivation from the policy together explain having a plan to quit smoking?_ 

The data was saved as an SPSS file, so you will have to use the foreign library and the read.spss command to open the file, like this:

```{r}
# open the UKY data set
library(foreign)
uky <- read.spss(file = "https://tinyurl.com/ydyns6e6", to.data.frame = T)
```

### Hacker edition

Complete the standard edition and add a graph of predicted probabilities like the one toward the end of this tutorial: https://stats.idre.ucla.edu/r/dae/logit-regression/. Your graph should predicted probabilities (y-axis) by percentage of friends who smoke (x-axis) with lines colored by motivation.

**UKY codebook**

The data are from a survey of University of Kentucky students following the implementation of a campus-wide smoking policy. The questions that were asked and the way the variables were coded are below:

- age - What is your age? (in years)
- age_initiation - How old were you when you smoked your first cigarette? (in years)
- gender - What is your sex? (0=female, 1=male)
- race - What is your race? (0=Non-white, 1=White)
- risky_behavior - Scale score indicating participation in risky activities (higher = riskier)
- quit - If you smoke, do you plan to quit? (0=No, 1=Yes)
- parent - Do you have a parent who smokes? (0=No, 1=Yes)
- sibling_smoke - Do you have a sibling who smokes? (0=No, 1=Yes)
- friends_smoke - Among your friends, what percentage do you think smoke?
- familiar_policy - How familiar are you with the rules and / regulations regarding the upcoming tobacco-free policy at the University of Kentucky? (0=unfamiliar, 1=familiar)
- policy_success - the policy will be successful with reducing secondhand smoke exposure and encouraging people to quit (scale from 1 to 9 where 1 is not at all and 9 is very successful)
- motivation - thinks the campus policy will increase my motivation to quit (0=No, 1=Yes)

